Recent Advances in Networking

### 2013-07-26 Compiled/chapter9/Content/conclusion.tex
fix typo

{\bf Top layer developments}. This type of solution aims at building a radically new system sitting on top of the legacy protocol stack, at the application layer. Essentially, such an approach considers the Internet as a black box providing a service equivalent to a cable connecting source(s) and destination(s), and provides novel mechanisms efficiently using this cable to cope with IP-disruptive characteristics. One example of such construction is the experimental Delay Tolerant Networking (DTN) architecture developed by the Internet Research Task Force (IRTF) \cite{DTNRG} \cite{ietf:rfc5050} \cite{ietf:rfc5326}, including a specific routing protocol targeting DTNs \cite{ietf:rfc6693}.\ \\ \ \\

>>>>

{\bf Top layer developments}. This type of solution aims at building a radically new system sitting on top of the legacy protocol stack, at the application layer. Essentially, such an approach considers the Internet as a black box providing a service equivalent to a cable connecting source(s) and destination(s), and provides novel mechanisms efficiently using this cable to cope with IP-disruptive characteristics. One example of such construction is the experimental Delay Tolerant Networking (DTN) architecture developed by the Internet Research Task Force (IRTF) \cite{DTNRG} \cite{rfc5050} \cite{rfc5326}, including a specific routing protocol targeting DTNs \cite{rfc6693}.\ \\ \ \\

### 2013-07-26 Compiled/chapter9/Content/specific.tex
fix typo

IETF design and standardization work in the reactive routing realm for mobile ad hoc networks first led to the Ad-hoc On-demand Distance Vector protocol (AODV) \cite{AODV-RFC3561}; the efforts in proactive routing, in turn, led to the Optimized Link State Routing (OLSR) \cite{OLSR-RFC3626}. A distance vector protocol, AODV operates in an \textit{on-demand} fashion, acquiring and maintaining routes only while needed for carrying data, by way of \textit{Route Request}-\textit{Route Reply} exchanges. A link state protocol, OLSR is based on periodic control messages exchanges, and each router proactively maintaining a routing table with entries for all destinations in the network. OLSR provides low delays in forwarding and has a predictable, constant control overhead -- at the expense of requiring memory in each router for maintaining complete network topology. AODV limits the memory required for routing state to that for actively used routes -- at the expense of delays for the \textit{Route Request}-\textit{Route Reply} exchange to take place, and control overhead dependent on data flows. \ \\ \ \\

@@

Based on the operational experience acquired through AODV and OLSRv1, the IETF is currently designing and developing successors for OLSR and AODV. In the first case, the IETF community involved in OLSR has standardized OLSR version 2 (OLSRv2) \cite{OLSRv2} and its related components (packet format \cite{packetbb-RFC5444} \cite{tlv-RFC5497}, NHDP \cite{NHDP-RFC6130}). Work on AODV version 2 \cite{ietf:aodvv2} has started, as AODV derivative flourished: IEEE 802.11s \cite{IEEE-802.11s-principles}, which is based on AODV, and the G3-PLC standard \cite{itu:PLC-G3}, published in 2011, which specifies the use of the {\em 6LoWPAN Ad hoc Routing Protocol} (LOAD, specified in {\tt draft-daniel-6lowpan-adhoc-routing}) \cite{ietf:I-D:LOAD} at the MAC layer, for providing layer 2 routing for utility (electricity) metering networks. 

@@

The IETF efforts on routing over 802.15.4 initially led to LOAD \cite{ietf:I-D:LOAD}, a derivative of AODV adapted for L2-addresses and mesh-under routing, and with some simplifications over AODV (\eg~removal of intermediate node replies and sequence numbers). However, 6LoWPAN was addressing other issues regarding adapting IPv6 for IEEE 802.15.4, such as IP packet header compression, and efforts to solve routing issues were suspended. In parallel with these efforts, the IETF has also specified the ``Routing Protocol for Low-power lossy networks'' (RPL), designed to support 6LoWPAN networks in a route-over configuration \cite{ietf:RFC6550:RPL, rpl-ipso}. \ \\ \ \\

@@

OLSR version 1 was standardized in RFC 3626 \cite{OLSR-RFC3626}. The work continues as OLSR version 2 (OLSRv2 \cite{OLSRv2}), which retains the same basic algorithms as its predecessor, however offers various improvements, {\em e.g.} a modular and flexible architecture allowing extensions, such as security, to be developed as add-ons to the basic protocol. \ \\ \ \\

@@

The Ad hoc On-Demand Distance Vector (AODV) \cite{AODV-RFC3561} protocol enables dynamic, self-starting, multi-hop routing between participating mobile  routers wishing to establish and maintain an ad hoc network.  AODV allows mobile nodes to obtain routes quickly for new destinations,  and does not require nodes to maintain routes to destinations that  are not in active communication. \ \\ \ \\

@@

AODV was initially standardized as an experimental RFC in 2003 \cite{AODV-RFC3561}. Derivatives of AODV include 802.11s (with HWMP \cite{IEEE-802.11s-principles}) and LOADng \cite{ietf:I-D:LOADng}. In the following of this section, the basic mechanisms of AODV, \textit{Route Discovery} and \textit{Route Maintenance} are explained. Then a main derivative work of AODV, called LOADng, is also introduced. Note that at the time of writing, work on AODVv2 \cite{ietf:aodvv2} has just started, and thus we will not elaborate further on AODVv2 in this chapter.

@@

The {\em Lightweight On-demand Ad hoc Distance-vector Routing Protocol - Next Generation} (LOADng) \cite{ietf:I-D:LOADng} is derived from AODV. Compared to AODV \cite{AODV-RFC3561}, it has more concise and flexible message format, and simplified message processing, which makes it more adapted to networks with constrained devices, such as sensor networks. It is also used for ITU Standard G. 9956 \cite{itu:PLC-G3}. \ \\ \ \\

@@

\item Different address lengths are supported -- from full 16 bytes IPv6 addresses over 6 bytes MAC addresses and 4 bytes IPv4 addresses to shorter 1 and 2 bytes addresses such as RFC 4944 \cite{ietf:rfc4944}.  The only requirement is, that within a given routing domain, all addresses are of the same address length.
\item Control messages are carried by way of the Generalized MANET Packet/Message Format \cite{packetbb-RFC5444}. 
\item  Using RFC 5444 \cite{packetbb-RFC5444}, control messages can include TLV (Type-Length-Value) elements, permitting protocol extensions to be developed.

@@

RPL -- the {\em Routing Protocol for Low Power and Lossy Networks}  \cite{ietf:RFC6550:RPL} -- is an IPv6 routing protocol designed and standardized by the ROLL Working Group in the IETF. It is is intended to be the IPv6 protocol for LLNs and sensor networks, applicable in all kinds of deployments and applications of LLNs.

@@

The basic construct in RPL is a ``Destination Oriented Directed Acyclic Graph'' (DODAG), depicted in Figure \ref{f:dodag} . In a converged LLN, each RPL router has identified a stable set of parents, each of which is a potential next-hop on a path towards the ``root'' of the DODAG, as well as a preferred parent. Each router, which is part of a DODAG (i.e. has selected parents) will emit DODAG Information Object (DIO) messages, using link-local multicast, indicating its respective rank in the DODAG (i.e. distance to the DODAG root according to some metric(s), in the simplest form hop-count). Upon having received a (number of such) DIO messages, a router will calculate its own rank such that it is greater than the rank of each of its parents, select a preferred parent and then itself start emitting DIO messages. The emission of DIO message is controlled by \textit{Trickle} Algorithm \cite{RFC6206-trickle}, to reduce the flooding overhead. \ \\ \ \\

>>>>

IETF design and standardization work in the reactive routing realm for mobile ad hoc networks first led to the Ad-hoc On-demand Distance Vector protocol (AODV) \cite{rfc3561}; the efforts in proactive routing, in turn, led to the Optimized Link State Routing (OLSR) \cite{rfc3626}. A distance vector protocol, AODV operates in an \textit{on-demand} fashion, acquiring and maintaining routes only while needed for carrying data, by way of \textit{Route Request}-\textit{Route Reply} exchanges. A link state protocol, OLSR is based on periodic control messages exchanges, and each router proactively maintaining a routing table with entries for all destinations in the network. OLSR provides low delays in forwarding and has a predictable, constant control overhead -- at the expense of requiring memory in each router for maintaining complete network topology. AODV limits the memory required for routing state to that for actively used routes -- at the expense of delays for the \textit{Route Request}-\textit{Route Reply} exchange to take place, and control overhead dependent on data flows. \ \\ \ \\

@@

Based on the operational experience acquired through AODV and OLSRv1, the IETF is currently designing and developing successors for OLSR and AODV. In the first case, the IETF community involved in OLSR has standardized OLSR version 2 (OLSRv2) \cite{OLSRv2} and its related components (packet format \cite{rfc5444} \cite{rfc5497}, NHDP \cite{rfc6130}). Work on AODV version 2 \cite{ietf:aodvv2} has started, as AODV derivative flourished: IEEE 802.11s \cite{IEEE-802.11s-principles}, which is based on AODV, and the G3-PLC standard \cite{itu:PLC-G3}, published in 2011, which specifies the use of the {\em 6LoWPAN Ad hoc Routing Protocol} (LOAD, specified in {\tt draft-daniel-6lowpan-adhoc-routing}) \cite{ietf:I-D:LOAD} at the MAC layer, for providing layer 2 routing for utility (electricity) metering networks. 

@@

The IETF efforts on routing over 802.15.4 initially led to LOAD \cite{ietf:I-D:LOAD}, a derivative of AODV adapted for L2-addresses and mesh-under routing, and with some simplifications over AODV (\eg~removal of intermediate node replies and sequence numbers). However, 6LoWPAN was addressing other issues regarding adapting IPv6 for IEEE 802.15.4, such as IP packet header compression, and efforts to solve routing issues were suspended. In parallel with these efforts, the IETF has also specified the ``Routing Protocol for Low-power lossy networks'' (RPL), designed to support 6LoWPAN networks in a route-over configuration \cite{rfc6550, rpl-ipso}. \ \\ \ \\

@@

OLSR version 1 was standardized in RFC 3626 \cite{rfc3626}. The work continues as OLSR version 2 (OLSRv2 \cite{OLSRv2}), which retains the same basic algorithms as its predecessor, however offers various improvements, {\em e.g.} a modular and flexible architecture allowing extensions, such as security, to be developed as add-ons to the basic protocol. \ \\ \ \\

@@

The Ad hoc On-Demand Distance Vector (AODV) \cite{rfc3561} protocol enables dynamic, self-starting, multi-hop routing between participating mobile  routers wishing to establish and maintain an ad hoc network.  AODV allows mobile nodes to obtain routes quickly for new destinations,  and does not require nodes to maintain routes to destinations that  are not in active communication. \ \\ \ \\

@@

AODV was initially standardized as an experimental RFC in 2003 \cite{rfc3561}. Derivatives of AODV include 802.11s (with HWMP \cite{IEEE-802.11s-principles}) and LOADng \cite{ietf:I-D:LOADng}. In the following of this section, the basic mechanisms of AODV, \textit{Route Discovery} and \textit{Route Maintenance} are explained. Then a main derivative work of AODV, called LOADng, is also introduced. Note that at the time of writing, work on AODVv2 \cite{ietf:aodvv2} has just started, and thus we will not elaborate further on AODVv2 in this chapter.

@@

The {\em Lightweight On-demand Ad hoc Distance-vector Routing Protocol - Next Generation} (LOADng) \cite{ietf:I-D:LOADng} is derived from AODV. Compared to AODV \cite{rfc3561}, it has more concise and flexible message format, and simplified message processing, which makes it more adapted to networks with constrained devices, such as sensor networks. It is also used for ITU Standard G. 9956 \cite{itu:PLC-G3}. \ \\ \ \\

@@

\item Different address lengths are supported -- from full 16 bytes IPv6 addresses over 6 bytes MAC addresses and 4 bytes IPv4 addresses to shorter 1 and 2 bytes addresses such as RFC 4944 \cite{rfc4944}.  The only requirement is, that within a given routing domain, all addresses are of the same address length.
\item Control messages are carried by way of the Generalized MANET Packet/Message Format \cite{rfc5444}. 
\item  Using RFC 5444 \cite{rfc5444}, control messages can include TLV (Type-Length-Value) elements, permitting protocol extensions to be developed.

@@

RPL -- the {\em Routing Protocol for Low Power and Lossy Networks}  \cite{rfc6550} -- is an IPv6 routing protocol designed and standardized by the ROLL Working Group in the IETF. It is is intended to be the IPv6 protocol for LLNs and sensor networks, applicable in all kinds of deployments and applications of LLNs.

@@

The basic construct in RPL is a ``Destination Oriented Directed Acyclic Graph'' (DODAG), depicted in Figure \ref{f:dodag} . In a converged LLN, each RPL router has identified a stable set of parents, each of which is a potential next-hop on a path towards the ``root'' of the DODAG, as well as a preferred parent. Each router, which is part of a DODAG (i.e. has selected parents) will emit DODAG Information Object (DIO) messages, using link-local multicast, indicating its respective rank in the DODAG (i.e. distance to the DODAG root according to some metric(s), in the simplest form hop-count). Upon having received a (number of such) DIO messages, a router will calculate its own rank such that it is greater than the rank of each of its parents, select a preferred parent and then itself start emitting DIO messages. The emission of DIO message is controlled by \textit{Trickle} Algorithm \cite{rfc6206}, to reduce the flooding overhead. \ \\ \ \\
